---
title: "Spanstatz"
subtitle    : Driving Sports Excellence Through Data Science
author      : "Sai S Sampathkumar"
logo        : SmallLogo.png
date: "14 July 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Our Vision


**To build a sustainable data driven sports eco-system in India by leveraging sports data science to enhance the ROI for owners, athletes, fans & communities
**

## Our Values

*We are TIRELESS*

- Transparency
- Integrity
- Respect 
- Empathy
- Loyalty 
- Empowerment
- Sustainability
- Social Impact


## Salary Comparison of US Leagues vs. Indian Premier League 2016-17


```{r ,cache=TRUE, eval = TRUE, results='hide'}
# call the necessary library packages
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(tidyr)))

# Loading MLB Data and creating MLB plot data set [source: wwww.sportrac.com]
mlb_contract_data <- read.csv("C:/Users/ssais/Documents/1. SPANSTATZ/8. Analysis/Datasets/MLB_Player_Salaries_2016_17.csv")
mlb_contract_data$leaguename <- rep("Major League Baseball",length = nrow(mlb_contract_data))
for (i in seq(along = mlb_contract_data$yearfreeagent))
  if (mlb_contract_data$yearfreeagent[i] == 0)
    mlb_contract_data$yearfreeagent[i] <- NA
for (i in seq(along = mlb_contract_data$age))
  if (mlb_contract_data$age[i] == 0)
    mlb_contract_data$age[i] <- NA
mlb_contract_data$salarymm <- mlb_contract_data$annualavg/1000000
mlb_contract_data$league <- rep("MLB",length = nrow(mlb_contract_data))
mlb_contract_data$salmmnorm <- (mlb_contract_data$salarymm - median(mlb_contract_data$salarymm))/ (max(mlb_contract_data$salarymm) - min(mlb_contract_data$salarymm))

print(summary(mlb_contract_data))
mlbdata_plot <- mlb_contract_data[,c("salarymm","leaguename","salmmnorm","league")]

#Loading NBA Data and creating NBA plot data set [source: wwww.sportrac.com]
nba_contract_data <- read.csv("C:/Users/ssais/Documents/1. SPANSTATZ/8. Analysis/Datasets/NBA_Player_Salaries_2016_17.csv")
nba_contract_data$leaguename <- rep("National Basketball Association",length = nrow(nba_contract_data))
for (i in seq(along = nba_contract_data$yearfreeagent))
  if (nba_contract_data$yearfreeagent[i] == 0)
    nba_contract_data$yearfreeagent[i] <- NA
for (i in seq(along = nba_contract_data$age))
  if (nba_contract_data$age[i] == 0)
    nba_contract_data$age[i] <- NA
nba_contract_data$salarymm <- nba_contract_data$annualavg/1000000
nba_contract_data$league <- rep("NBA",length = nrow(nba_contract_data))
nba_contract_data$salmmnorm <- (nba_contract_data$salarymm - median(nba_contract_data$salarymm))/ (max(nba_contract_data$salarymm) - min(nba_contract_data$salarymm))

print(summary(nba_contract_data))
nbadata_plot <- nba_contract_data[,c("salarymm","leaguename","salmmnorm","league")]


#Loading NFL Data and creating NFL plot data set [source: wwww.sportrac.com]

nfl_contract_data <- read.csv("C:/Users/ssais/Documents/1. SPANSTATZ/8. Analysis/Datasets/NFL_Player_Salaries_2016_17.csv")
nfl_contract_data$leaguename <- rep("National Football League",length = nrow(nfl_contract_data))
for (i in seq(along = nfl_contract_data$yearfreeagent))
  if (nfl_contract_data$yearfreeagent[i] == 0)
    nfl_contract_data$yearfreeagent[i] <- NA
for (i in seq(along = nfl_contract_data$age))
  if (nfl_contract_data$age[i] == 0)
    nfl_contract_data$age[i] <- NA
nfl_contract_data$salarymm <- nfl_contract_data$annualavg/1000000
nfl_contract_data$league <- rep("NFL",length = nrow(nfl_contract_data))
nfl_contract_data$salmmnorm <- (nfl_contract_data$salarymm - median(nfl_contract_data$salarymm))/ (max(nfl_contract_data$salarymm) - min(nfl_contract_data$salarymm))

print(summary(nfl_contract_data))
nfldata_plot <- nfl_contract_data[,c("salarymm","leaguename","salmmnorm","league")]

#Loading MLB Data and creating MLB plot data set [source: http://www.cricmetric.com/ipl/salary/]
ipl_contract_data <- read.csv("C:/Users/ssais/Documents/1. SPANSTATZ/8. Analysis/Datasets/IPL_Player_Salaries_2017.csv")
ipl_contract_data$leaguename <- rep("Indian Premier League",length = nrow(ipl_contract_data))

ipl_contract_data$salarymm <- ipl_contract_data$Salary_USD/1000000
ipl_contract_data$league <- rep("IPL",length = nrow(ipl_contract_data))
ipl_contract_data$salmmnorm <- (ipl_contract_data$salarymm - median(ipl_contract_data$salarymm))/ (max(ipl_contract_data$salarymm) - min(ipl_contract_data$salarymm)) 
print(summary(ipl_contract_data))
ipldata_plot <- ipl_contract_data[,c("salarymm","leaguename","salmmnorm","league")]


# Merging data for plot
plotting_data_frame <-rbind(ipldata_plot,mlbdata_plot, nbadata_plot, nfldata_plot)
cutpoints <- quantile(plotting_data_frame$salarymm, probs = c(0.25,0.50,0.75,1.0), na.rm = TRUE)
plotting_data_frame$Salcat <- cut(plotting_data_frame$salarymm,cutpoints, labels = c("2_MED", "3_HIGH", "4_ULHI"))
plotting_data_frame$Salcat <- as.character(plotting_data_frame$Salcat)
x <- length(plotting_data_frame$salarymm)
#converting NA's in Q1 to LOW
NAS <- which(is.na(plotting_data_frame$Salcat))

for (i in 1:length(NAS)){
    if (is.na(plotting_data_frame$Salcat[NAS[i]])) 
        { plotting_data_frame$Salcat[NAS[i]]<-"1_LOW"}}

# Removing data
rm("ipl_contract_data")
rm("mlb_contract_data")
rm("nba_contract_data")
rm("nfl_contract_data")
rm("ipldata_plot")
rm("mlbdata_plot")
rm("nbadata_plot")
rm("nfldata_plot")


```


```{r ,cache = TRUE, eval= TRUE, results='asis', fig.align='center', message=FALSE, fig.height=1.5, fig.width=4}

# Dataset "Plotting_data_frame" is what we will use to generate plots
# Plot salary densities 

suppressMessages(suppressWarnings(library(plotly)))

# Histogram of total_bill, divided by sex and smoker

g <- ggplot(plotting_data_frame, aes(x=salarymm, fill=Salcat)) +
    geom_histogram(position="identity", alpha=0.4)+
    facet_grid(. ~ league)
ggplotly(g)
```

## Contact Us

We would love to hear from you:

### sss@spanstatz.com
